Fusing Filters with Integer Linear Programming

### 2014-05-15 2014betterfusionforfilters/Main.tex
Add TODO to intro
Cut back related work section
condense abstract

@@

\begin{abstract}
Minimising memory traffic is key to compiling functional, collection oriented array programs into efficient code. To achieve this, it is not sufficient to just fuse subsequent array operations into a single, more complex computation; we also have to \emph{cluster} separate traversals of the same array into one single loop. Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused. However, these approaches have one serious restriction: they can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities. The technique we present in this paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP solver to find good clusterings.

%The original abstract:
%Recent work on array fusion shows how to extract data-flow graphs from functional programs and compile them into efficient imperative loops. However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more. Prior work shows how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle operations like filter, which produce arrays of a different size than their input. We extend the existing ILP approach with support for size changing operators, using an external ILP solver to find good clusterings.
\end{abstract}

@@

\input{section/01-Introduction.tex}
\input{section/02-CNF.tex}
\input{section/03-SizeInference.tex}
\input{section/04-ILP.tex}
\input{section/05-Benchmarks.tex}
\input{section/06-Related.tex}
\section{Conclusion}

>>>>

% TODO: Fix up writing to use the words combinator and operator consistently
% We seem to be using "combinator" for the functions in the original program, 
% but "operator" for to name the abstract operation that combinator performs.

% \TODO{Cite work on clock inference for data flow languages. Eg http://www.irisa.fr/prive/talpin/papers/hldvt05.pdf}

@@

\begin{abstract}
Minimising memory traffic is key to compiling functional, collection oriented array programs into efficient code.
It is not sufficient to simply fuse subsequent array operations into a single computation; we also have to cluster \emph{separate} traversals of the same array into a single loop.
Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused.
However, these approaches can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities.
This paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP solver to find good clusterings.
\end{abstract}

@@

\input{section/01-Introduction.tex}
\input{section/02-CNF.tex}
\input{section/03-SizeInference.tex}
\input{section/04-ILP.tex}
\input{section/05-Benchmarks.tex}
\input{section/06-Related.tex}

### 2014-05-15 2014betterfusionforfilters/section/01-Introduction.tex
condense abstract

%!TEX root = ../Main.tex
\section{Introduction}
\label{s:Introduction}
A collection-oriented programming model is well suited for expressing data-parallel computations, as it exposes the communication patterns to the compiler without the need for complicated loop analysis.  It is essential, however, that the compiler combines those operations and compiles them into a small number of efficient imperative loops as a naive, operation-by-operation translation, would lead to high memory consumption and memory traffic, and result in poor performance. This is a well known problem, and techniques to fuse subsequent array operations into a single, more complex one, have been presented in \cite{coutts2007streamfusion, gao1993collective, kennedy1993typedfusion, kennedy2001fastgreedy, keller2010repa}, to name just a few. Fusion is, in general, not sufficient to minimise memory traffic: first, we often need to re-arrange the array operations and cluster those that iterate over the same structures. As shown in  (\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find such a clustering. Unfortunately, since they cannot handle operations like filter, which produce arrays whose size differs from the input arrays.  The technique we present in this paper can handle both multi-loop fragments as well as size-altering operations.

To compile subgraphs which are the result of our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}. It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa} as it 1) fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.

>>>>

%!TEX root = ../Main.tex
\section{Introduction}
\label{s:Introduction}
A collection-oriented programming model is well suited for expressing data-parallel computations, as it exposes the communication patterns to the compiler without the need for complicated loop analysis.  It is essential, however, that the compiler combines those operations and compiles them into a small number of efficient imperative loops as a naive, operation-by-operation translation, would lead to high memory consumption and memory traffic, and result in poor performance. This is a well known problem, and techniques to fuse subsequent array operations into a single, more complex one, have been presented in \cite{coutts2007streamfusion, gao1993collective, kennedy1993typedfusion, kennedy2001fastgreedy, keller2010repa}, to name just a few. Fusion is, in general, not sufficient to minimise memory traffic: first, we often need to re-arrange the array operations and cluster those that iterate over the same structures. As shown in (Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find such a clustering. Unfortunately, they cannot handle operations like filter, which produce arrays whose size differs from the input arrays.  The technique we present in this paper can handle both multi-loop fragments as well as size-altering operations.

To compile subgraphs which are the result of our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}. It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa}) as it 1) fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
